---
title: "Briggs Journal Club 01/18/21"
author: "sbsambado"
date: "1/18/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Today's Goals

+ 1. Create a github
+ 2. Create a github repo
+ 3. Clone a repo into a new R proj
+ 4. Go over basics of Rproj and .Rmd
+ 5. Tidyverse examples (NPS)
      +  find sbsambado repo 206
      +  find data file 'National Parks Visit'
      +  download .csv directly into your R proj
      +  go along with code
  
**Examples taken from Prof Allison Horst at Bren class 206 or 244**

# National Parks

```{r load libraries and data for NP}
library(readr)
library(tidyverse)
library(ggplot2)

np_visits <- read_csv("Desktop/National Parks Visitation Data.csv")
#View(np_visits)
```


```{r explore data}

names(np_visits) # notice the messy column names 
dim(np_visits) # 18 variables, 21560 observations
class(np_visits) # show difference between read.csv vs read_csv
head(np_visits)
tail(np_visits)

# help gauge numbers in your data set, explore your outcome of interest

#what was the max/min number of visitors per year

# could make individual vectors (don't reccommend)
MaxVis <- max(np_visits$Visitors, na.rm = TRUE)
MinVis <- min(np_visits$Visitors, na.rm = TRUE)

# or just do summary function (recommend)
summary(np_visits$Visitors)

```

```{r data wrangling}

df1 <- select(np_visits, State:YearRaw) # only select columns from state through yearraw, store as new data frame df1
#View(df1)

df2 <- filter(df1, State == "CA" & Type == "National Park" & YearRaw >= 1950) # filter to retain rows that are CA and NP
#View(df2)

df3 <- arrange(df2, Code, YearRaw)# arrange in alphabetical order in Code, THEN by ascending value in Year Raw
#View(df3)

df4 <- mutate(df3, kVis = Visitors/1000) # add new column to df3, ocnverts to units of 1000s of visitors
#View(df4) # 578 observations, 7 variables

summary(df4) # like to reevaluate my dataset after cleaning it up; YearRaw = character?
unique(np_visits$YearRaw)

df5 <- filter(df4, YearRaw != "Total") # remove 'total' values in the YearRaw column (!= does NOT match)

summary(df5) # we need to change year to numeric vs character

df5$YearRaw <- as.numeric(df5$YearRaw)

class(df5$YearRaw) # make sure it changes
```

Intro to piping
```{r intro to piping}

# %>% (command + shift + m) is saying "and then..."

utah_np <- np_visits %>% 
  filter(State == "UT", Type == "National Park") %>% 
  select(Name, Visitors, YearRaw) %>% 
  filter(YearRaw != "Total") %>% 
  mutate(mill_vis = Visitors/1000000)

utah_np # 403 observations, 4 variables (make sure they understand where the 4th column came in since we selected for only 3 variables)

# have people suggest another state example
ca_np <- np_visits %>% 
  filter(State == "CA",  Type == "National Park") %>% 
  select(Name, Visitors, YearRaw) %>% 
  filter(YearRaw != "Total") %>% 
  mutate(mill_vis = Visitors/1000000)

ca_np # 789 observations, 4 variables
  
  
```


```{r graph it}

# ask for structure of ggplot
# ggplot(dataframe, aes(x = variable1, y = variable2))+
# geom()

# when i'm building my plot i do it piece by piece and then assign variable name
visitor_graph <- 
  ggplot(df5, aes(x = YearRaw, y = kVis))+
  geom_point(aes(color = Code)) +
  geom_line(aes(color = Code)) +
  ggtitle("CA NP Visitations\n1950-2016") +
  xlab("\nYear") + # \n just adds a blank line before this label
  ylab("Thousands of Visitors\n") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(color = "black")) +
  scale_x_continuous(breaks = c(1950, 1960, 1970, 1980, 1990, 2000, 2010, 2020))
 visitor_graph 

VisBoxplot <- 
  ggplot(df5, aes(x = Code, y = kVis)) +
  geom_boxplot(aes(fill = Code)) +
  theme_bw() +
  ggtitle("CA NP Visits") +
  xlab("NP") +
  ylab("number of visitors") +
  scale_x_discrete(breaks = c("CHIS","DEVA","JOTR","KICA","LAVO","PINN","REDW","SEQU","YOSE"), 
                   labels = c("Channel Islands","Death Valley","Joshua Tree","Kings Canyon","Lassen Volcanic","Pinnacles","Redwoods","Sequoia","Yosemite")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = .65))

VisBoxplot 
```

If there's time for more

# Fish

Tidyverse examples (salmon)
  a. find sbsambado repo ESM244
  b. find folder called data
  c. find data file 'noaa_fisheries.csv'
  c. download .csv directly into your R proj
  d. go along with code
  
```{r load libraries and data for FISH}

library(tidyverse)
library(here)
library(janitor)
library(kableExtra)

us_landings <- read_csv("Desktop/noaa_fisheries.csv")
#View(us_landings)

```

## Break data into ways that make sense
We'll break this up into two pieces:

1. A tidier version of the entire data frame
2. A subset that only contains salmon information
```{r split dataset}

# tidy whole dataframe
landings_tidy <- us_landings %>% 
  clean_names() %>% # make column headers snake_case
  mutate(
    state = str_to_lower(state),
    afs_name = str_to_lower(afs_name)
  ) %>% #converts character columns to lowercase
  mutate(dollars_num = parse_number(dollars_usd))

# just get salmon
salmon_landings <- landings_tidy %>% 
  mutate(afs_clean = str_remove(afs_name, pattern = "aggregate")) %>% 
  filter(str_detect(afs_clean, pattern = "salmon")) %>%  #detect matches for crab
  separate(afs_clean, into = c("group", "species"), sep = ",")

```


## Find grouped summary data

Find the annual total US landings and dollar value (summing across all states) for each type of salmon using `group_by()` + `summarize()`:
```{r}
salmon_summary <- salmon_landings %>% 
  group_by(year, species) %>% 
  summarize(
    tot_landings = sum(landings_pounds),
    tot_value = sum(dollars_num)
  )


salmon_landings_graph <- ggplot(salmon_summary, aes(x = year, y = tot_landings, group = species)) +
  geom_line(aes(color = species)) +
  theme_bw() +
  labs(x = "year", y = "US commercial salmon landings")

salmon_landings_graph
#ggsave(plot = salmon_landings_graph, here("figures", "us_salmon_landings.png"),
      # height = 5, width = 8)
```


## 2015 commercial fisheries value by state

Now, let's create a finalized table of the top 5 states (by total commercial fisheries value) for 2015 .

Remember that we already created a tidied data frame, **landings_tidy**. 

*Critical thinking for data wrangling workflow: Why does it make sense (especially now) that we broke our previous wrangling into two steps before getting our salmon subset?* 

```{r}
state_value <- landings_tidy %>% 
  filter(year %in% c(2015)) %>% 
  group_by(state) %>% 
  summarize(
    state_value = sum(dollars_num, na.rm = TRUE),
    state_landings = sum(landings_pounds, na.rm = TRUE)
  ) %>% 
  arrange(-state_value) %>% 
  head(5)
```

What are some ways we'd want to finalize it for a report?

- Change units (to millions of dollars and pounds)
- Update column names
- Capitalize the state names

## Making a nice HTML table

finalized data frame
```{r}
state_table <- state_value %>% 
  mutate('Fisheries value ($ millions)' = round(state_value / 1e6, 2),
         'Landings (million pounds)' = round(state_landings / 1e6, 1)) %>% 
  select(-state_value, - state_landings) %>% 
  rename(State = state) %>% 
  mutate(State = str_to_title(State))
```

Now, use `kable()` + `kableExtra` to nicely format it for HTML:

```{r}
kable(state_table) %>% 
  kable_styling(bootstrap_options = "striped",
                full_width = FALSE) %>% 
  add_header_above(c("", "2015 US commercial fisheries - top 5 states by value" = 2))
```