---
title: "TidyverseTutorial_Briggslab_2020"
author: "sbsambado"
date: "12/8/2020"
output:
  html_document: default
  pdf_document: default
---

This file is intended to be an intro to tidyverse for the undergraduates in the Briggs lab.

Here are helpful links I derive a lot of my information about tidyverse & `dplyr` : 

+ Basic Tidyverse Concepts
https://homerhanumat.github.io/r-notes/tidyverse-concepts.html

+ Tidyverse with animations
https://www.garrickadenbuie.com/project/tidyexplain/

+ Tidyverse-cookbook
https://rstudio-education.github.io/tidyverse-cookbook/tidy.html

+ YaRrr! The pirate’s guide to R **chapter 10.4**
(I also think this is an excellent guide for beginners to R!)
https://bookdown.org/ndphillips/YaRrr/dplyr.html 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# necessary libraries
library(tidyverse)
library("devtools")
library(mosaicData)

# two packages to accompany the text
devtools::install_github("homerhanumat/bcscr")
devtools::install_github("homerhanumat/tigerData")
```

## *tidyverse isn't a package, it's a collection of packages* 
^^^ just like universities are made up of colleges that have specific functions

# Tidy data has the following rules

+ 1. Each **variable** has its own **column**

+ 2. Each **observation** has its own **row**

+ 3. Each **value** has its own **cell**

# Basics

**1. pipe operator `%>` : chains function together**
```{r}

# using pipe operator
"hello" %>% rep(times = 4)

# without pipe operator
rep("hello", times = 4)

```

**2. tibble : type of dataframe suited for tidyverse**
~ good for large datasets
```{r}

# dataframe
class(bcscr::m111survey)


# convert to tibble
survey <- as_tibble(bcscr::m111survey)
```


**3. dplyr package: tool for data manipulation** 

+ a. `filter()` and `select()` for sub-setting

+ b. `mutate()` for transforming variables

+ c. `group_by()` and `summarise()` for numerical summaries


a. examples of **`filter()` and `select()`** to chose optimal observations & variables
```{r}

# usually filter() and select() are your first arguments when you get a dataset

survey %>% # call tibble
  filter((sex == "male" & height > 70) | (sex == "female" & height < 55)) %>% # filter picks out rows/observations ## we're selecting observations that are men > 70 in tall AND women < 55 in tall
  select(sex, height, fastest) # selects picks out columns/variables


# let's leave out columns/variables ; use `-`

survey %>%
  select(-ideal_ht, -love_first) # still have lots of columns left!

# want to see how many observations you have left post subsetting?

survey %>%
  filter(love_first == "yes" & fastest > 120) %>%
  nrow() # count the rows/observations left

```

b. examples of **`mutate()`** to transform variables
```{r}
# keep new variable name on LEFT side of `=`
# keep function on RIGHT side of `=` that depends on the current value in the tibble

survey %>%
  mutate(dareDevil = fastest > 125) %>% # make new column/variable called dareDevil ## this will provide TRUE (> 125) /FALSE (< 125) for each observation with regards to its fastest value
  select(sex, fastest, dareDevil) # select desired variables

# You can double dip mutations within one line of code

survey %>%
  mutate(dareDevil = fastest > 125, 
         speedKmHR = fastest * 1.60934) %>% # made two new variables/columns based on previous data in column 'fastest')
  ggplot(aes(x = dareDevil, y = GPA)) + # plot these new variable values
  geom_boxplot(fill = 'burlywood', outlier.alpha = 0) + # using boxplot to visualize datat
  geom_jitter(width = 0.2) # avoid points overlapping and making them a certain size

  

```



c. examples of **`group_by()` and `summarise()`** to generate numerical summaries 

+ *^^ my most used functions in dplyr*

```{r}

CPS85 %>% # tibble that comes from the `mosaicData`
  group_by(sex) %>% # divide observations based on their values/observations in the variable 
  summarize(meanWage = mean(wage)) # make new tibble that has mean wage for each level of sex, you will only have two columns/variables in this tibble 1) sex and 2) meanWage

# create more than 1 summary variable

CPS85 %>% 
  group_by(sex) %>% 
  summarize(meanWage = mean(wage),
            n = n()) # n() is the count function, very helpful for most summaries ## will count total observations for your variable meanWage


```

**other summaries you can do are:**

+ minimum value
+ first quartile (q1)
+ median 
+ third quartile (q3)
+ max

if you want to run all 5 summaries (which you should), use `fivenum()` function
```{r}

# summaries for entire tibble (men and women included)
CPS85 %>%
  .$wage %>% # select one column/vector of information you want to summarize
  fivenum() # run all 5 summary statistics

# summaries broken down for each sex (more appropriate of comparisons!)

CPS85 %>%
  group_by(sex) %>% 
  summarise(n = n(), # count observations
            min = fivenum(wage)[1],
            Q1 = fivenum(wage)[2],
            median = fivenum(wage)[3],
            Q3 = fivenum(wage)[4],
            max = fivenum(wage)[5])

# you can also compare multiple variables by group
CPS85 %>% 
  group_by(sector, sex) %>%  # looking at SEX and SECTOR
  summarise(n = n(),
            min = fivenum(wage)[1],
            Q1 = fivenum(wage)[2],
            median = fivenum(wage)[3],
            Q3 = fivenum(wage)[4],
            max = fivenum(wage)[5])
```


# More practice!

```{r}
data("flights", package = "nycflights13")
```

**filter out NA cases**

+ Want to correctly count number of flights that left airport
```{r}
flights %>% 
  filter(!is.na(dep_delay)) %>% 
  group_by(origin) %>% 
  summarise(departures = n(),
            meanDelay = mean(dep_delay))
```

**filter for a specific date**

+ Find flights that occurred on June 26
```{r}
flights %>% 
  filter(month == 6 & day == 26) %>% 
  ggplot(aes(x = origin, y = distance)) +
    geom_violin(fill = "burlywood") +
    geom_jitter(width = 0.25, size = 0.1)
```

**filter for specific date, variable value and types of variables**

+ Find flights that occurred on June 26 that were over 4000 mi and chose only relevant information of origin, destination, and distance
```{r}
flights %>% 
  filter(month == 6 & day == 26 & distance > 4000) %>% 
  select(origin, dest, distance)
```
  
  