---
title: "DANC Workshop on reproducible & publishable figures"
author: "sbsambado"
date: "11/16/2021"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

## The "purpose"

This workshop is intended for the participants in Data Analysis Coding (DAnC) club at UC Santa Barbara. The goal of this workshop is to **learn some techniques** on how to make publishable figures & tables.

## Some blah, blah, blah

More often than not, when I attend live coding events most things don't stick with me. When I was first learning R, this realization got me really frazzled, however once I changed my perspective that these live coding events were great opportunities to **1) ask blunt questions** about each line of code I didn't get & **2) keep these rich resources** of .Rmd files that I could revisit when I had my own data. Not to mention how great it is to meet other R enthusiasts!

To be honest, **I don't intend for us to get through all of these examples today** or maybe these examples won't be useful to you for some time. However I hope when you are scrambling for time and have to give a quick presentation or submit a report for a class/ research lab/ job opportunity, you can refer to this template and save yourself some time & sanity. 

What I have here **comes entirely from the internet** and kindness of Stack Overflow strangers. I have never made up anything novel in R. I try to cite my sources as best as can but honestly a lot of this comes from random google searches like 'how to move a legend box in ggplot' and dumb things I have learned 34985 previous times. 

## Actual content list

1. Plotting **plots** - let's keep building those ggplot skills with a familiar data set

2. Quick n dirty **summary** tables

3. Nice-ish **model output** tables 

4. **Google satellite maps** - this is something I recently did and it took forever to make. I hope when you get to this coding stage you will not have to struggle as much as I did. You could make these things in PowerPoint but I was determined to find a solely R method. 


### The packages and data needed for this workshop! 

(with the exception of section 4)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)

#### neccessary libraries

# install.packages("ggplot2")
library(ggplot2) # for plot making 

# install.packages("tidyverse")
library(tidyverse) # for data wrangling

# install.packages("ggpubr")
library(ggpubr)

remotes::install_github("allisonhorst/palmerpenguins") # install package from Prof Horst's Github
library(palmerpenguins) # call package into library

# install.packages("gtsummary")
library(gtsummary)

# install.packages("lme4")
library(lme4)

# install.packages("effects")
library(effects)

# install.packages("jtools")
library(jtools)


#### neccessary data

# dataset for section 1
penguins <- palmerpenguins::penguins # more info: https://allisonhorst.github.io/palmerpenguins/

# dataset for section 2
trial <-gtsummary::trial # more info: https://www.danieldsjoberg.com/gtsummary/reference/trial.html

# dataset for section 3
penguins <- palmerpenguins::penguins # more info: https://allisonhorst.github.io/palmerpenguins/

# dataset for section 4
uc <- read.csv("uc_campuses.csv") # this is a separate dataset that is sent out with the material

```

### 1. More ggplots

I think the main takeaway from this section is pay attention to common functions/arguments such as:

+ **theme()** : to me, this is my most powerful tool while plotting. Spend some time on the ?theme() help page to better understand  all the cool visual things you can do with it

+ **labs()** : don't be that person whose lab titles make no sense and throw some units in while you're there

+ **element_text()** : this is where you define the size and font type (i.e. italic or bold)

+ color = "green" gives a green *outline* to geom

+ fill = "green" give s geom that is *filled in* by green



**A. The set up** 

```{r}

## necessary packages and data for this section

library(ggplot2) # to make ggplots

remotes::install_github("allisonhorst/palmerpenguins") # install package from Prof Horst's Github
library(palmerpenguins) # call package into library
penguins <- palmerpenguins::penguins # give dataset a name to work with 



## data checking & refining

# check out what kind of data you're working with
str(penguins) # 344 rowsobservations, 8 columns/groupings

# Later I will want to have labels be capitalized in my graph, this is one way of doing it
supp.labs <- c("Female", "Male") # state your new labels 
names(supp.labs) <- c("female", "male") # override the old lowercase labels with uppercase labels

```

**B. How it started**

```{r}

## most basic ggplot at it's core

# plot 1
ggplot(data = na.omit(penguins), 
       aes(x = flipper_length_mm, y = bill_depth_mm)) +
  geom_point()

# plot 2 
ggplot(na.omit(penguins), aes(x = as.factor(year), y = body_mass_g, group_by = sex)) +
  geom_boxplot()
  
```

**C. How it's going (as of now)**

```{r}
(plot1 <- ggplot(data = na.omit(penguins), # omit NAs from dataset for visualization purposes
       aes(x = flipper_length_mm, y = bill_depth_mm, color = species)) + # choose our variables of interest
  geom_point() + # geom_point makes sense since there's 2 numeric values
  facet_grid(island ~ sex, # want to create a "grid" for our geom_points. facet_wrap is another option for 1 variable
             scale = "free_y",  # want to y axis to best fit each group, not for entire dataset
             labeller = labeller(sex = supp.labs)) + # want to change labels from lower to uppercase
  scale_color_manual(values = c("#0073C2FF", "#EFC000FF", "#CD534CFF")) + # choose colors manually, could do "red","blue",etc ( make sure use scale_COLOR not scale_FILL since in line 66 we said COLOR = species)
  labs(x = "Flipper Length (mm)", # change the x axis labels to something nicer to look at
       y = "Bill Depth (mm)", # change the y axis labels to something nicer to look at
       color = "Species") + # this changes the title of your legend
  theme_bw() + # this sets the entire theme to a white background with light bordering - play around with theme_*()
  theme(axis.title = element_text(face = "bold"), # gives us bold text for axes labels Body & Flipper
        strip.background = element_rect(fill = "gray95"), # changes the colors of the facet_grid rectangles
        strip.text = element_text(size = 10, face = "bold"), # change size of font in facet_grid label and make it bold
        legend.position = "bottom", # move legend to bottom of graph (there are many ways of doing this, this is one)
        legend.text = element_text(size = 10), # make legend text same size as facet_grid text
        legend.title = element_text(face = "bold"))) # make legend title a bold font


(plot2 <- ggplot(na.omit(penguins), # omit NAs from dataset for visualization purposes
                 aes(x = as.factor(year), # i did as.factor() bc if you don't year will show up as 2017.5 which doesn't make sense in terms of this team's sampling. try it out for yourself to see without as.factor()
                     y = body_mass_g, group_by = sex)) + # 2 other variables of interest
  geom_violin(aes(fill = sex), trim = TRUE, draw_quantiles = c(0.25, 0.5, 0.75)) + # geom_violin shows the distribution of your points, trim gives it a sharper look, drew quantiles for each violin
  facet_wrap(~species) + # breaking up data by 1 variable of interest
  labs(x = "Year", # change the x axis labels to something nicer to look at
       y = "Body Mass (g)", # change the y axis labels to something nicer to look at
       fill = "Sex") + # this changes the title of your legend.. notice how I used FILL instead of COLOR
  theme_bw() + # this sets the entire theme to a white background with light bordering - play around with theme_*()
  theme(axis.title = element_text(face = "bold"), # same as above exmple
        strip.background = element_rect(fill = "gray95"), # same as above example
        strip.text = element_text(size = 10, face = "bold"), # same as above example
        legend.position = c(.92, .14), # horizontal, verticle # i manually placed my legend instead of using a command, pros is that you can creatively fit it around your data, cons is that it takes a little fiddling
        legend.box.background = element_rect(color = "black"), # giving a rectangle box with a black outline to my legend
        legend.text = element_text(size = 10), # make legend text same size as facet_grid text
        legend.title = element_text(face = "bold"))) # make legend title a bold font


```

**D. One simple ggpubr() example**

sorry I advertised this as a ggpubr() workshop...

```{r}
# state which comparisons you would like to make between the different levels of your factor `species`

my_comparisons <- list(c("Adelie", "Chinstrap"), c("Adelie", "Gentoo"), c("Chinstrap", "Gentoo"))

# slightly different format that ggplot() but relatively similar concepts
ggboxplot(na.omit(penguins), x = "species", y = "body_mass_g",  # data you want to plot
         add = "boxplot", add.params = list(fill = "grey89")) + # fill in boxplots
  stat_compare_means(comparisons = my_comparisons, aes(label = "p.signif")) + # print p-value of your comparisons
  stat_compare_means(label.y = 7500, label.x = .7) + # print statistical test used to make comparisons
  labs(x = "Species", # change axes labels
       y = "Body Mass (g)")

```


### 2. Statistical summary tables

A core concept in this section in the pipe ( %>% ). Basically whenever you see **%>%**, in your head say **'and then do this'**. It's just notation that tries to speed up the coding process by building on previous lines instead of starting over.

*Notes:*
+ categorical : data that can only take on one *characteristic* (ie, dog or cat, can't be cat-dog, not *that* cat-dog)

+ continuous : data can take on any *numeric* value (ie, 1.2, 33, .00000002)


**A. The set up**

```{r}
# once again here are the necessary bits for this section
library(gt)
library(gtsummary)
trial <-gtsummary::trial

# more info: https://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html

```

**B. How it started**

```{r}
# basic
trial %>%
  tbl_summary()

dim(trial) # 200 rows/observations, 8 columns/groupings

# that's a lot of info, let's focus on a subset of variables

## so if my question was how is treatment type impacted by age, grade of tumor, and chemical marker?

trial_subset <- trial %>%
  select(trt, age, grade, marker)

dim(trial_subset) #200 rows/observations, 3 columns/groupings

```

**C. How it's going**

```{r}

# i'm going to build this up during live coding so you can see how we go from step to step, but the final product will look something like this


trial_table <- trial_subset %>% # use the subset of data that is to our interest
  tbl_summary( # whatever content we want in the table, put it inside the tbl_summary() function
    by = trt, # we are most interested in the effect of treatment 
    statistic = list( # let's add some stats
                     all_categorical() ~ "{n}/{N} ({p}%)", # I want all categorical values to show sampled/total (percentage)
                     all_continuous() ~ "{mean} ({sd})"), # I want all categorical values to show mean & standard deviation
    missing_text = "(Missing)", # instead of unknown I want them called missing -- things happen
                     digits = marker ~ 3, # round only my marker variable to 3 digits (you could specify all_* if you'd like)
    label = grade ~ "Tumor Grade") %>% # let's give this label something more digestable
  add_p() %>% # let's add the wretched p-value
  add_stat_label(location = "column") # let's let people know what stats we're showing, although could omit this
```

**D. Let's make it fancy**

```{r}

# let's build off our our 'trial_table' from the previous section which has all the INFO we want, but now let's give it the LOOK we want

(trial_table_fancy<-trial_table %>%
  modify_header(label ~ "**Variable**")%>% # make header columns bold
  modify_caption("**Table 1. Patient Characteristics**") %>%# give the table a name
  bold_labels() %>% # bold labels
  modify_spanning_header(c("stat_1","stat_2") ~ "**Drug Type**") )# make a sub header
  


```

### 3. Model outputs

**A. The set up**

```{r}
library(lme4)
library(effects)
library(jtools)
library(palmerpenguins) 


penguins <- palmerpenguins::penguins # more info: https://allisonhorst.github.io/palmerpenguins/

# more info: https://jtools.jacob-long.com/articles/summ.html
```

**B. Make the models**

these may not be the "best" statistical models but it'll do for the purposes of this workshop

```{r}

penguins
mod1 <- lm(body_mass_g ~ species, data = penguins)
mod2 <- lm(body_mass_g ~ island, data = penguins)
mod3 <- lm(body_mass_g ~ bill_depth_mm, data = penguins)
mod4 <- lm(body_mass_g ~ sex, data = penguins)

```

**C. Visualize estimates for all models**

```{r}
plot_summs(mod1, mod2, mod3, mod4, # models you want to compare
           scale = TRUE, # scale for all covariates
           model.names = c("Species", "Island", "Bill Depth (mm)", "Sex"), # give models specific names
           inner_ci_level = .8, # thickness of the ci band
           coefs = c("Chinstrap" = "speciesChinstrap",
                     "Gentoo" = "speciesGentoo",
                     "Dream Island" = "islandDream", 
                     "Torgersen Island" = "islandTorgersen",
                     "Bill Depth (mm)" = "bill_depth_mm",
                     "Sex" = "sex"))
#ggsave("Coefficients.jpeg", dpi = 320)
```

**D. Print out pdf of model outpus**

```{r}
export_summs(mod1, mod2, mod3, mod4,
           scale = TRUE,
           error_format = "{statistic}, 95% CI [{conf.low}, {conf.high}]",
           borders = .5,
  outer_borders = 0,
             to.file = "docx",  # could also do pdf
             file.name = "Modeloutputs.docx") # sent it to docx so I could manually edit it there but usually if you have less models you can fit on one page of a pdf. just takes some fiddling with some of the parameters with export_summs()
```


### 4. Google Satellite Maps

**A. The set up**

```{r}
# I didn't include these packages in the very beginning because some of these packages are quite annoying to install... so only do this if you're ready to explore google maps


# this might be an overkill but usually this combo of packages get me to my final result
library(ggmap)
library(readr)
library(tidyverse)
library(ggrepel)
library(ggtext)

# i think these may not be necessary but just in case
library(sf)
library(raster)
library(spData)
library(spDataLarge)

## Google Earth is being extra and making individual computers get APIs before using their data

# some links that helped me figure this out

# set up API: https://developers.google.com/earth-engine/reference/rest

# get API to connect to your R studio: https://jstnbraaten.medium.com/running-google-earth-engine-from-rstudio-cloud-3faba86af8a5

```

**B. Call in data & clean it up**

```{r}
# load data
uc <- read.csv("uc_campuses.csv")

# make sure it is in the lat/lon format
names(uc)

# define location
location <- with(uc,
                 c(lon = median(lon), # make sure your column names are lon
                   lat = median(lat)))
```

**C. How it started**

```{r}

# get map with appropriate zoom

uc_map <- get_map(location = location, 
                  zoom = 6, # bigger number gives you smaller zoom
                  source = "google", # this can be "stamen"
                  maptype = "satellite") # terain also looks nice

ggmap(uc_map) # put your map into a ggmap format
```

**D. How it's going**

```{r}

ggmap(uc_map) + # base map
  geom_point(data= uc ,aes(x = lon, y = lat, color = students), size = 5) + # points
    geom_label_repel( # labels
      aes(x = lon, y = lat, label = campus), data = uc,
      size = 4, # size of circles
      box.padding = .6,  # how big boxes are
      max.overlaps = Inf,  
      direction = "both",
      xlim = c(-Inf,Inf),
      min.segment.length = 0, #every box gets a line 
      segment.size = 1) + # line width
  coord_cartesian(clip = "off") + # fit all the boxes in the coordinates
  labs(x = "Longitude", y = "Latitude", color = "Enrolled Students") + # change labels 
    theme(axis.title = element_text(face = "bold")) 
 # ggsave("reserve.png") 
```

